---
title: "CB2 improves power of cell detection in droplet-based single-cell RNA sequencing data"
authors: "Zijian Ni and Christina Kendziorski"
package: scCB2
date: 06/23/2020
output: 
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{CB2 improves power of cell detection in droplet-based single-cell RNA sequencing data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}  
---


```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```


# Introduction

Droplet-based single-cell RNA sequencing (scRNA-seq) is a powerful and widely-used approach for profiling genome-wide gene expression 
in individual cells. Current commercial droplet-based technologies such as 10X Genomics utilize gel beads, each containing 
oligonucleotide indexes made up of bead-specific barcodes combined with unique molecular identifiers (UMIs) and oligo-dT tags 
to prime polyadenylated RNA. Single cells of interest are combined with reagents in one channel of a microfluidic chip, 
and gel beads in another, to form gel-beads in emulsion, or GEMs. Oligonucleotide indexes bind polyadenylated RNA within each GEM 
reaction vesicle before gel beads are dissolved releasing the bound oligos into solution for reverse transcription. 
By design, each resulting cDNA molecule contains a UMI and a GEM-specific barcode that, ideally,
tags mRNA from an individual cell, but this is often not the case in practice.  To distinguish true cells from background barcodes in
droplet-based single cell RNA-seq experiments, we introduce **CB2** and `scCB2`, its corresponding R package.

**CB2** extends the EmptyDrops approach by introducing a clustering step that groups similar barcodes and 
then conducts a statistical test to identify groups with expression distributions that vary from the background.
While advantages are expected in many settings, users will benefit from 
noting that **CB2** does not test for doublets or multiplets and, consequently, some of the high count identifications 
may consist of two or more cells. Methods for identifying multiplets may prove useful after applying **CB2**. 
It is also important to note that any method for distinguishing cells from background barcodes is technically correct in 
identifying low-quality cells given that damaged cells exhibit expression profiles that differ from the background. 
Specifically, mitochondrial gene expression is often high in damaged cells.
Such cells are typically not of interest in downstream analysis and should therefore be removed. 
The GetCellMat function in `scCB2` may be used toward this end.

# Quick Start

## Installation

Install from Bioconductor:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("scCB2")

```

## All-in-one function

`QuickCB2` is an all-in-one function to apply CB2 on 10x Cell Ranger raw data and get a matrix of real cells identified by CB2 under default settings. By specifying `AsSeurat = TRUE`, a Seurat object is returned so that users can directly apply the Seurat pipeline for downstream analyses.

Usage: 
```{r, eval=FALSE}
library(scCB2)

# If raw data has three separate files within one directory
# and you want to control FDR at the default 1%:
RealCell <-  QuickCB2(dir = "/path/to/raw/data/directory")

# If raw data is in HDF5 format and 
# you'd like a Seurat object under default FDR threshold:
RealCell_S <-  QuickCB2(h5file = "/path/to/raw/data/HDF5", 
                        AsSeurat = TRUE)
```

An example illustrating how it works and what the final output looks like can be found at the end of *Detailed Steps*.

## Running Speed

The computational speed is related to the size and structure of input datasets. As a reference, scCB2 running on a medium-size dataset (around 30,000 genes and 8,000 cells after cell calling) using 6 cores takes less than 10 minutes.

## Saving a sparse matrix to 10x format

For users who would like to save the CB2 output cell matrix to 10x format (e.g. "matrix.mtx", "barcodes.tsv" and "genes.tsv"), there are existing packages to help. For example in package `DropletUtils`:

```{r, eval=FALSE}
DropletUtils::write10xCounts(path = "/path/to/save/data",
                             x = RealCell)
```

# Detailed Steps

## Read count matrix from 10x output raw data

Currently, the most widely-used droplet-based protocol is 10x Chromium. Our package provides functions to directly read 10x Cell Ranger output files and generate a feature-by-barcode count matrix that may be read into R. Public 10x datasets can be found [here](https://www.10xgenomics.com/resources/datasets). 

Our package contains a small subset of 10x data, `mbrainSub`, corresponding to the first 50,000 barcodes of [1k Brain Cells from an E18 Mouse](http://cf.10xgenomics.com/samples/cell-exp/2.1.0/neurons_900/neurons_900_raw_gene_bc_matrices.tar.gz).

We first generate 10x output files of `mbrainSub`, then read it using our built-in functions.

```{r}
library(scCB2)
library(SummarizedExperiment)

data(mbrainSub)

data.dir <- file.path(tempdir(),"CB2_example")
DropletUtils::write10xCounts(data.dir,
                             mbrainSub,
                             version = "3")

list.files(data.dir)
```

For Cell Ranger version <3, the raw data from 10x Cell Ranger output contains "barcodes.tsv", "genes.tsv" and "matrix.mtx". For Cell Ranger version >=3, the output files are "barcodes.tsv.gz", "features.tsv.gz" and "matrix.mtx.gz". We now read these files back into R and compare with original data matrix.

```{r}
mbrainSub_2 <- Read10xRaw(data.dir)
identical(mbrainSub, mbrainSub_2)
```

If raw data is not from the 10x Chromium pipeline, a user may manually create the feature-by-barcode count matrix with rows representing genes and columns representing barcodes. Gene and barcode IDs should be unique. The format of the count matrix can be either a sparse matrix or standard matrix.

## Run **CB2** to distinguish real cells from empty droplets

The main function `CB2FindCell` takes a raw count matrix as input and returns real cells, test statistics, and p-values. Now we apply `CB2FindCell` on `mbrainSub`, controlling FDR at 0.01 level (Default), assuming all barcodes with total count less than or equal to 100 are background empty droplets (Default), using 2 cores parallel computation (Default: number of total cores in the machine minus 2). For detailed information, see `?CB2FindCell`.

```{r}
CBOut <- CB2FindCell(mbrainSub, Ncores = 2)
str(assay(CBOut)) # cell matrix
str(metadata(CBOut)) # test statistics, p-values, etc
```

## Extract real cell matrix

If readers are not interested in the output testing information, `GetCellMat` can extract the real cell matrix directly from `CB2FindCell` output. It also provides a filtering option to remove broken cells based on the proportion of mitochondrial gene expressions. Now we apply `GetCellMat` on `CBOut`, filtering out cells whose mitochondrial proportions are greater than 0.25 (Default: 1, No filtering).

```{r}
RealCell <- GetCellMat(CBOut, MTfilter = 0.25)
str(RealCell)
```

## Downstream analysis

After `CB2` pre-processing, the real cell matrix is still in matrix format, so it can be directly used in downstream statistical analyses. For example, if we want to use the *Seurat* pipeline, we can easily create a Seurat object using

```{r}
SeuratObj <- Seurat::CreateSeuratObject(counts = RealCell, 
                                        project = "mbrain_example")
SeuratObj
```

## All-in-one function

Under default parameters, we can directly use the all-in-one function `QuickCB2` to get the real cell matrix from 10x raw data. 

```{r}
RealCell_Quick <- QuickCB2(dir = data.dir, Ncores = 2)
str(RealCell_Quick)
```

Now it's ready for downstream analysis such as normalization and clustering. Example Seurat tutorial: https://satijalab.org/seurat/v3.1/pbmc3k_tutorial.html

# Session Information
```{r}
sessionInfo()
```

# Citation

Ni, Z., Chen, S., Brown, J., & Kendziorski, C. (2020). CB2 improves power of cell detection in droplet-based single-cell RNA sequencing data. Genome Biology, 21(1), 137. https://doi.org/10.1186/s13059-020-02054-8
